tidyverse for data wrangling, gridExtra for arranging multiple plots, ggplot for plotting.

```{r}
library("tidyverse")
library("gridExtra")
library("ggplot2")
library("plotly")
```

---
Hackathon
Monday, February 24

Dataset: World Happiness Report Datasets (2015-2019)
---

Grab data for each year. Most of the data processing here should be done with for-loop and list, but wasn't
completely sure on how to do it properly. So each year's dataset is processed one by one separately.

```{r}
df_2015 <- read.csv("/Users/lubis/Documents/study/dataanalysis/hackathon/world-happiness/2015.csv", header=TRUE)
df_2016 <- read.csv("/Users/lubis/Documents/study/dataanalysis/hackathon/world-happiness/2016.csv", header=TRUE)
df_2017 <- read.csv("/Users/lubis/Documents/study/dataanalysis/hackathon/world-happiness/2017.csv", header=TRUE)
df_2018 <- read.csv("/Users/lubis/Documents/study/dataanalysis/hackathon/world-happiness/2018.csv", header=TRUE)
df_2019 <- read.csv("/Users/lubis/Documents/study/dataanalysis/hackathon/world-happiness/2019.csv", header=TRUE)
```

Check types of each year to work only on the common ones.

```{r}
str(df_2015)
str(df_2016)
str(df_2017)
str(df_2018)
str(df_2019)
```

Since the feature names change in some of the years, we specific on which features to work, and renaming the features
correspondingly.
1. Country.or.region renamed into Country
2. Happiness.Rank renamed into Overall.rank
3. Happiness.Score renamed into Score
4. Economy..GDP.per.Capita. renamed into GDP.per.capita
5. Health..Life.Expectancy. renamed itno Healthy.life.expectancy
6. Freedom renamed into Freedom.to.make.life.choices
7. Trust..Government.Corruption renamed into Perceptions.of.corruption

For 2015, aside form renaming some features, 3 features (Standard.Error, Family, Dystopia.Residual) are removed.

```{r}
df_2015_rename <- df_2015 %>%
  rename(Overall.rank = Happiness.Rank, Score = Happiness.Score, GDP.per.capita = Economy..GDP.per.Capita.,
         Healthy.life.expectancy = Health..Life.Expectancy., Perceptions.of.corruption = Trust..Government.Corruption.) %>%
  select(-c(Standard.Error, Family, Dystopia.Residual))
str(df_2015_rename)
```

2016 is treated the same way as 2015.

```{r}
df_2016_rename <- df_2016 %>%
  rename(Overall.rank = Happiness.Rank, Score = Happiness.Score, GDP.per.capita = Economy..GDP.per.Capita.,
         Healthy.life.expectancy = Health..Life.Expectancy., Perceptions.of.corruption = Trust..Government.Corruption.) %>%
  select(-c(Lower.Confidence.Interval, Upper.Confidence.Interval, Family, Dystopia.Residual))
str(df_2016_rename)
```

For 2017, 4 features (Whisker.high, Whisker.low, Family, Dystopia.Residual) are removed.

```{r}
df_2017_rename <- df_2017 %>%
  rename(Overall.rank = Happiness.Rank, Score = Happiness.Score, GDP.per.capita = Economy..GDP.per.Capita.,
         Healthy.life.expectancy = Health..Life.Expectancy., Perceptions.of.corruption = Trust..Government.Corruption.) %>%
  select(-c(Whisker.high, Whisker.low, Family, Dystopia.Residual))
str(df_2017_rename)
```

2018 and 2019 go through similar process. Some features are renamed, and 1 feature (Social.support) is removed
from the DataFrame.

```{r}
df_2018_rename <- df_2018 %>%
  rename(Country = Country.or.region, Freedom = Freedom.to.make.life.choices) %>%
  select(-c(Social.support))
df_2019_rename <- df_2019 %>%
    rename(Country = Country.or.region, Freedom = Freedom.to.make.life.choices) %>%
    select(-c(Social.support))
str(df_2018_rename)
str(df_2019_rename)
```

All of these renamed DataFrames are put in new variables with "_rename" attached to their name.

------
TAKE REGION AND PLUG TO THE REMAINING YEARS

The feature Region which denotes the region of each country (North America, Western Europe, etc.) is missing from 2017.
So this needs to be added to the later datasets (2017, 2018, 2019)
------

```{r}
plug_region <- df_2015 %>% select(Country, Region)
df_2017_rename <- left_join(df_2017_rename, plug_region)
df_2018_rename <- left_join(df_2018_rename, plug_region)
df_2019_rename <- left_join(df_2019_rename, plug_region)
```

Add new feature year for each dataset.

```{r}
df_2015_rename$year <- 2015
df_2016_rename$year <- 2016
df_2017_rename$year <- 2017
df_2018_rename$year <- 2018
df_2019_rename$year <- 2019
```


```{r}
View(df_2015_rename)
View(df_2016_rename)
View(df_2017_rename)
View(df_2018_rename)
View(df_2019_rename)
```

Bind rows is just stacking everything by rows, distinguished by year.

```{r}
df <- bind_rows(df_2015_rename, df_2016_rename, df_2017_rename, df_2019_rename)
df
```

Then plot using plotly.

```{r}
p <- ggplot(df, aes(Healthy.life.expectancy, Score, color = Region)) +
  geom_point(alpha=0.5, aes(size = GDP.per.capita, frame = year, ids = Country)) +
  scale_x_log10() + xlim(0.05, 1.00)

p <- ggplotly(p)

p
```

