---
title: "Let's make it functional part 1: R series"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**2ndMeeting - Saturday, November 23, 2019**

**Tutors and Organizers:** 
- Pande Putu Erawijantati
- Felix Salim
- Mia Fitria Utami

## Content:
  - Data types and structure (review)
	- Loop
	- Functions
	- Analysing multiple datasets
Sources: 
- Composing function
 <http://swcarpentry.github.io/r-novice-inflammation/02-func-R/index.html>
- Application
 <http://swcarpentry.github.io/r-novice-inflammation/03-loops-R/index.html>
- Problem set:
TBA

*Notes: the notes was created in Python 3 kernel with extension for runing R code because some of the participants already familiar with Python.

### Data types and structure

R contains 6 data types and structures, which are:
* character
* numeric (real or decimal)
* integer
* logical
* complex

Please read and try the example here: <https://swcarpentry.github.io/r-novice-inflammation/13-supp-data-structures/>

### Functions
ref: <http://swcarpentry.github.io/r-novice-inflammation/02-func-R/index.html>

Functions is a defined sections of program that perform a specific task. 
Defining a functions is very usefull if you are going to perform similar task for several times. 
For example, you have several data that you will visualize as a barplot, then it is better to create a defined function to plot a barplot.
In the example of [Creating functions](http://swcarpentry.github.io/r-novice-inflammation/02-func-R/index.html). 
Below is the example to convert Fahrenheit into Celsius.

Defining the function. Pay attention to the syntax of defining the functions in R and Python below.

**R**
```{r}
fahrenheit_to_celsius <- function(temp_F) {
  temp_C <- (temp_F - 32) * 5 / 9
  return(temp_C)
}
fahrenheit_to_celsius(100)
```

**Python**
```{python}
import numpy as np
def fahrenheit_to_celcius(temp_F):
  temp_C = ((np.float(temp_F)-32)*5/9)
  return(temp_C)
fahrenheit_to_celcius(100)
```

Now, let's try our functions understanding in analysing multiple set of data.
We will use the same set of data that we use in the [fist meeting](http://swcarpentry.github.io/r-novice-inflammation/data/r-novice-inflammation-data.zip)
Previously, we only try to analyze the `inflammation-01.csv`. We still have 12 data to go. 
Let's try to apply functions to analyse our inflammation data.
Step by step tutorial will be following the example in <http://swcarpentry.github.io/r-novice-inflammation/03-loops-R/index.html>
First please make sure your are working in the right directory as we studied in the previous meeting.
Next, Let's create a function called analyze that creates graphs of the minimum, average, and maximum daily inflammation rates for a single data set. Apply that function to the `inflammation-01.csv`
```{r}
analyze <- function(filename) {
  # Plots the average, min, and max inflammation over time.
  # Input is character string of a csv file.
  dat <- read.csv(file = filename, header = FALSE)
  avg_day_inflammation <- apply(dat, 2, mean)
  plot(avg_day_inflammation)
  max_day_inflammation <- apply(dat, 2, max)
  plot(max_day_inflammation)
  min_day_inflammation <- apply(dat, 2, min)
  plot(min_day_inflammation)
}
#call function
analyze("data/inflammation-01.csv")
```
What about using the same function to analyse the second dataset?
```{r}
analyze("data/inflammation-02.csv")
```
One line do the magic thing right? 
Hmmm, but this is not the easier way to do that. Let's do some magic `Loops` to replicate the same action of one function to all of 12 data that we have.

## For Loops
To replicate same thing for several times, we can use `For Loops`. 
Suppose we want to print each word in a sentence. One way is to use six print statements:

**Bad example first**
```{r}
best_practice <- c("Let", "the", "computer", "do", "the", "work")
print_words <- function(sentence) {
  print(sentence[1])
  print(sentence[2])
  print(sentence[3])
  print(sentence[4])
  print(sentence[5])
  print(sentence[6])
}

print_words(best_practice)
```
Seems works, what's wrong?
Try this
```{r}
print_words(best_practice[-6])
```
R seems nice because it dose not throw an error, in python, this is a big NO.
Example in python.
```{python}
best_practice=["Let", "the", "computer", "do", "the", "work"] #this is call `list` in python, we will talk about this later.
def print_words(sentence):
  print(sentence[0]) #remember python start with 0 but R start with 1
  print(sentence[1])
  print(sentence[2])
  print(sentence[3])
  print(sentence[4])
  print(sentence[5])

print_words(best_practice)
```
```{python error=TRUE}
print_words(best_practice[-6])
```
Now, let's try using `For loops` instead and see how it works.
```{r}
print_words <- function(sentence) {
  for (word in sentence) {
    print(word)
  }
}

print_words(best_practice)
```
Please try anothe example in <http://swcarpentry.github.io/r-novice-inflammation/03-loops-R/index.html> before we are going to process our multiple files.

## Processing Multiple Files
Sometimes we do not need to create function for specific task if it is available in the programing languange. For example, to begin the analysis on multiple files, we have to list what files we are going to analyze. We do not need to write it ourselves because R already has a function to do this called `list.files.` If we run the function without any arguments, `list.files()`, it returns every file in the current working directory. We can understand this result by reading the help file `(?list.files)`. The first argument, path, is the path to the directory to be searched, and it has the default value of `"."` ("." is shorthand for the current working directory). The second argument, pattern, is the pattern being searched, and it has the default value of NULL. Since no pattern is specified to filter the files, all files are returned.
```{r}
list.files(path = "data", pattern = "csv")
```
```{r}
list.files(path = "data", pattern = "inflammation")
```
> Remember that, if you will be working with larger project, it is better to specified sub-directories for better organizations. Refer to [A quick guide to organizing computational biology projects](http://swcarpentry.github.io/r-novice-inflammation/03-loops-R/index.html) for more advice.

To show the complete path in directory you can use the argument `full.names = TRUE`
```{r}
list.files(path = "data", pattern = "csv", full.names = TRUE)
```

Let’s test out running our analyze function by using it on the first three files in the vector returned by `list.files`:

```{r}
filenames <- list.files(path = "data",  
                        # Now follows a regular expression that matches:
                        pattern = "inflammation-[0-9]{2}.csv",
                        #          |            |        the standard file extension of comma-separated values
                        #          |            the variable parts (two digits, each between 0 and 9)
                        #          the static part of the filenames
                        full.names = TRUE)
filenames <- filenames[1:3]
for (f in filenames) {
  print(f)
  analyze(f)
}
```
> In this lesson we saw how to use a simple for loop to repeat an operation. As you progress with R, you will learn that there are multiple ways to accomplish this. Sometimes the choice of one method over another is more a matter of personal style, but other times it can have consequences for the speed of your code. For instruction on best practices, see this [software carpentry supplementary lesson](http://swcarpentry.github.io/r-novice-inflammation/15-supp-loops-in-depth/) that demonstrates how to properly repeat operations in R.

## More challenge
See notebooks here, it also cover our next meeting topics.
<https://www.kaggle.com/rtatman/getting-started-in-r-load-data-into-r>
