---
title: "Answer for problem in Let's make it functional part 1: R series"
output: github_document
---
**Problem 1**

- What is the class of the object "surveys" ?
Answer: 
```{r}
surveys <- read.csv("data_raw/portal_data_joined.csv")
class(surveys)
```
- How many rows and how many columns are in this object?
```{r}
dim(surveys)
```
- How many species have been recorded during these surveys?
```{r}
length(unique(surveys$species))
```

**Problem 2**
1. Create a `data.frame` (surveys_200) containing only the data in row 200 of the surveys dataset.
```{r}
surveys_200 <- surveys[200, ]
surveys_200
```
2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
  - Use that number to pull out just that last row in the data frame.
  - Compare that with what you see as the last row using tail() to make sure it’s meeting expectations.
  - Pull out that last row using nrow() instead of the row number.
  - Create a new data frame (surveys_last) from that last row.
```{r}
last_row <- nrow(surveys)
surveys_last <- surveys[last_row, ]
surveys_last
```
3. Use nrow() to extract the row that is in the middle of the data frame. Store the content of this row in an object named surveys_middle.
```{r}
surveys_middle <- surveys[c(last_row / 2 - 1, last_row / 2) , ]
surveys_middle
```
4. Combine nrow() with the - notation above to reproduce the behavior of head(surveys), keeping just the first through 6th rows of the surveys dataset.
```{r}
surveys[-c(7:last_row), ]
```
Alternative to show the table, you can use `View()` which will show the table in new tab in Rstudio.




