---
title: "6thSemiHackathon_Session_Era"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Datasets

I will explore the trend from the world happiness datasets (2015-2019) specifically for Indonesia and compared it to the all region in South East Asia.
 
 
**Data preparation**

1. Prepare the directory

```{r}
dir.create("data_raw")
```

2. Download data from [kaggle-world happinness datasets](https://www.kaggle.com/unsdsn/world-happiness).

#example: (Notes: better download directly)
```
download.file(url="https://www.kaggle.com/unsdsn/world-happiness/download/pNRoj8OUY1j8qhSQbRnv%2Fversions%2FO5X6zWeGu2XCpwGL2QkE%2Ffiles%2F2015.csv?datasetVersionNumber=2",
              destfile = "data_raw/2015.csv")
```

3. Load all data

```{r}
library(tidyverse)
data_2015 = read_csv("data_raw/2015.csv")
data_2016 = read_csv("data_raw/2016.csv")
data_2017 = read_csv("data_raw/2017.csv")
data_2018 = read_csv("data_raw/2018.csv")
data_2019 = read_csv("data_raw/2019.csv")

```


4. Extract only Southeastern asia, modify columns name

```{r}
#Filter if region is Southeastern Asia
data_2015_AS <-data_2015 %>% 
    filter(Region=="Southeastern Asia") %>% 
    mutate(year=2015) %>% 
  rename_all(gsub, pattern = '[ ()]', replacement = '.') %>% 
  rename( GDP.per.Capita= Economy..GDP.per.Capita.) %>% 
  rename( Life.Expectancy= Health..Life.Expectancy.) %>% 
  rename( GDP.per.capita= GDP.per.Capita) %>% 
  rename(Perceptions.of.corruption = Trust..Government.Corruption.)

data_2016_AS <-data_2016 %>% 
  filter(Region=="Southeastern Asia") %>% 
  mutate(year=2016) %>% 
  rename_all(gsub, pattern = '[ ()]', replacement = '.') %>% 
  rename( GDP.per.Capita= Economy..GDP.per.Capita.) %>% 
  rename( Life.Expectancy= Health..Life.Expectancy.) %>% 
  rename( GDP.per.capita= GDP.per.Capita) %>% 
  rename(Perceptions.of.corruption = Trust..Government.Corruption.)

row_reg <- data_2015_AS$Country
```


```{r}
#add the region information to the 2017,2018,2019

data_2017_AS <-data_2017 %>% filter(Country %in% row_reg) %>% 
  mutate(year=2017,Region="Southeastern Asia") %>% 
  rename_all(gsub, pattern = '[ ()]', replacement = '.') %>% 
  rename( GDP.per.Capita= Economy..GDP.per.Capita.) %>% 
  rename( Life.Expectancy= Health..Life.Expectancy.) %>% 
  rename( GDP.per.capita= GDP.per.Capita) %>% 
  rename(Perceptions.of.corruption = Trust..Government.Corruption.)
  
data_2018_AS <-data_2018 %>% filter(`Country or region` %in% row_reg) %>% 
  mutate(year=2018,Region="Southeastern Asia") %>% 
  rename(Country = `Country or region`) %>% 
  rename_all(gsub, pattern = '[ ()]', replacement = '.') %>% 
  rename( Life.Expectancy= Healthy.life.expectancy) %>% 
  rename(Happiness.Score = Score) %>% 
  rename(Freedom = Freedom.to.make.life.choices)
  

data_2019_AS <-data_2019 %>% filter(`Country or region` %in% row_reg) %>% 
  mutate(year=2019,Region="Southeastern Asia") %>% 
  rename(Country = `Country or region`) %>% 
  rename_all(gsub, pattern = '[ ()]', replacement = '.') %>% 
  rename( Life.Expectancy= Healthy.life.expectancy) %>% 
  rename(Happiness.Score = Score) %>% 
  rename(Freedom = Freedom.to.make.life.choices)
```


5. Find intesect of dataframe rows

```{r}
#intersect
common <- reduce(list(colnames(data_2015_AS), colnames(data_2016_AS), colnames(data_2017_AS),colnames(data_2018_AS), colnames(data_2019_AS)), intersect)
```

6. Combine tables (naive ways)

```{r}
#select similar columns
dt_2015_AS <- data_2015_AS %>% select(one_of(common))
dt_2016_AS <- data_2016_AS %>% select(one_of(common))
dt_2017_AS <- data_2017_AS %>% select(one_of(common))
dt_2018_AS <- data_2018_AS %>% select(one_of(common)) %>% mutate(Perceptions.of.corruption = as.double(Perceptions.of.corruption))
dt_2019_AS <- data_2019_AS %>% select(one_of(common))

#combine
df_AS <- reduce(list(dt_2015_AS,dt_2016_AS,dt_2017_AS,dt_2018_AS,dt_2019_AS), bind_rows)
```

7. Plot trend from the data

a. Happiness Score

```{r}
ggplot(data = df_AS, mapping = aes(x = year, y = Happiness.Score)) +
  geom_line() +
facet_wrap(facets = vars(Country))
```

b. Generous

```{r}
ggplot(data = df_AS, mapping = aes(x = year, y = Generosity)) +
  geom_line() +
facet_wrap(facets = vars(Country))
```

What kind analysis we can go from here?

