---
title: "Answer for challenge in Data is beautiful part 1 - Hello Tidyverse! : R Seriesr"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation
```{r}

library(tidyverse)
surveys <- read_csv("data_raw/portal_data_joined.csv")

surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex
```


## Challenge 1

> To see the shape of distributions, replace the boxplot with a violin plot; see `geom_violin()`. 

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_violin(alpha = 0)
```

> 2. Sometimes, we need to change the scale for easier observations. Try to represent the weight on the log10 scale; see `scale_y_log10()`. 

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y =weight)) +
  geom_violin(alpha = 0,  draw_quantiles=0.5) + #draw quantiles to see the median
  scale_y_log10()
```

> 3. Create a boxplot for `hindfoot_length` and overlay the boxplot with jitter for actual measurements

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y =hindfoot_length)) +
  geom_violin(alpha = 0)
```

> 4. Add color for each data point based on which plot the sample was taken (`plot_id`)

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y =hindfoot_length, color=as.factor(plot_id))) +
  geom_boxplot(alpha = 0)
```

Let's put the color as a jitter instead.

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.1, aes(color = as.factor(plot_id))) +
  geom_boxplot(alpha = 0)
```



## Challenge 2

> Use what you just learned to create a plot that depicts how the average weight of each species changes through the years.

```{r}
yearly_weight <- surveys_complete %>%
                group_by(year, species_id) %>%
                 summarize(avg_weight = mean(weight))
ggplot(data = yearly_weight, mapping = aes(x=year, y=avg_weight)) +
   geom_line() +
   facet_wrap(vars(species_id)) +
   theme_bw()
```

